---
title: "UMD Fees"
---

Follow along with the demo to learn how to calculate percentage change using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.


### Task 1: Load the tidyverse


```{r}
library(tidyverse)
```




### Task 2: Load the UMD Fees dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv

```{r}
umd_fees <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/umd_fees.csv")
```

### Task 3: Calculate the percentage change in each category between Fall 2024 and Fall 2021 using `mutate()`

```{r}
umd_fees_pct <- umd_fees |> mutate(pct_change = (`Fall 2024`- `Fall 2021`)/`Fall 2021` * 100)
```


```{r}
umd_fees_pct_2524 <- umd_fees |> mutate(pct_change = (`Fall 2025`- `Fall 2024`)/`Fall 2024` * 100)

```


### Task 4: Describe our results

Write a few sentences about what you found in the data. What's the story here?

Comparing the student fees from fall 2025 to fall 2024, students are paying more money in total in fall 2025. The rate each fee increases are all different however. The highest increasing fee is student activities at 14.6%, while athletics and student sustainability do not increase. The other fees are increasing at different rates, for some reason. The story here is why are these fees increasing, are students seeing any benefits from these increased fees they have to pay.

tesitng character


```{r}
comp <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/class_composition.csv")

comp |>
  summarize(
    mean_first = mean(first_year),
    mean_soph = mean(sophomore),
    mean_junior = mean(junior),
    mean_senior = mean(senior),
    mean_grad = mean(grad)
  )
```


When you are done, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
