---
title: "final_project"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
```



This will be most of my data, I will have to manually input some of the rankings for 2021 because that year had less data than the rest. The only other data I need to collect is to find the same data of acceptance rates for peer universities to Maryland. I will select 10 other peer universities out of this data to do that.

Links to the sources: https://www.usmd.edu/IRIS/DataJournal/Applications/?report=UG-Applicants-First-Time-Admission&Residency=OUT#
https://www.usmd.edu/IRIS/DataJournal/Applications/?report=UG-Applicants-First-Time-Admission#
https://www.usmd.edu/IRIS/DataJournal/Applications/?report=UG-Applicants-First-Time-Admission&Residency=MD#
https://today.umd.edu/umd-rises-no-19-us-news-rankings-top-public-colleges-dd2ee090-44a7-4534-a0fa-903f46c8e9bf
https://github.com/frishberg/Archive-of-US-News-College-Rankings/blob/main/data.csv
https://www.aronfrishberg.com/projects/usnews
https://www.washington.edu/research/wp-content/uploads/US-NewsRankings_2021Undergraduate-2022Graduate_2021GlobalFinal.pdf



## The Data

```{r load-data}
umd_total <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTKW1tniQ6YzHPGymOLiG6FVWNttdRldiqiRGbfjNmu4D5ktljJSoVUQBZZgSSXowAQfxBZL2ZPgHw6/pub?output=csv")

umd_outstate <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTwL-MKQaRGvRLdNoyJdvy2iiVIJrURsu4b9PP4_hv7X3d5fGmK8XftphpWlKWdNlrXdVBjM3ua100K/pub?output=csv")

umd_instate <-read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQVATeqEyiCDPYh3sRUqSaNzK7T0CvRBXWN3TxsvOjjps3kspyODWp59geqYOKQrbVuVtQrCdPO65Fu/pub?output=csv")

us_news_rankings <-read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTMyKnrvJ48Xmj_Bebg806Z3yXWI3ahQvAn2YbxYsSuW0H1k1avXQTQo1cnBbiUvqUSQbzr6K6UYq1-/pub?output=csv")

```



## The Type of Test

The type of test I want to run for my project is a null hypothesis test. I want to pit Maryland versus 10 other similar Public Institutions and my Hypothesis is that Maryland's school ranking has risen higher than other schools, due to the change in admission rates changing. This will be similar to the HypothesisTesting assignment we did in class. I will try to find if the data proves my hypothesis, if it is statistically significant, meaningfully significant, and the effect size. 


Another test I can run is a Z-Score test similar to the CarTheft Z-score activity. This would be helpful to find if Maryland's change in rankings is close to the mean of school's change in rankings over time or if Maryland is an outlier compared to similar Universities.







