---
title: "HW11: Education Level and Survey Weighting"
author: "Your Name Here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

When a poll or survey gets published, the results almost always include a note about sample size and methodology. One thing you'll often see: "results have been weighted to reflect the U.S. adult population." But what does weighting actually *do* to the numbers — and does it matter for your story?

In this assignment, you'll compare weighted and unweighted results from a 2020 survey on voting intentions, focusing on how education level relates to whether someone planned to vote.

## The Data

```{r load-data}
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")
glimpse(nonvoters_data)
```

Key variables:

- `weight`: Each respondent's survey weight. A weight of 1.0 means they count as exactly one person. A weight of 2.0 means their response counts double (they represent an undersampled group). A weight of 0.5 means their response counts half as much (they represent an oversampled group).
- `Q21`: Voting intention — 1 = Yes, 2 = No, 3 = Unsure
- `educ`: Education level — College, Some college, High school or less

## Task 1: Examine the Weights (4 points)

Before comparing weighted and unweighted results, let's understand what the weights look like.

```{r examine-weights}
summary(nonvoters_data$weight)

ggplot(nonvoters_data, aes(x = weight)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Survey Weights",
    x = "Weight Value",
    y = "Number of Respondents"
  )
```

**Question 1:** Most weights cluster near 1.0, but some are much higher or lower. What does a weight of 0.2 tell you about that respondent's group in the sample? What about a weight of 3.5? Why might a pollster assign such extreme weights?

The weight being 0.2 tells us that the respondent's group was over represented in the sample. The pollster lowered their value so that they can count as 1/5th of a person in the weighted results. A weight of 3.5 is pretty much the complete opposite and the respondent's group was under sampled. This means then the pollster increased their value so much so that the final dataset can reflect the actual populaiton. Pollsters assign extreme weights to groups that either answer at a very low or high rate, and then weight the answers to balance the data out.

## Task 2: Education Distribution in the Sample (3 points)

```{r education-distribution}
nonvoters_data |>
  count(educ) |>
  mutate(percentage = round(n / sum(n) * 100, 1))
```

**Question 2:** College graduates are overrepresented in most surveys. Does that appear to be true here? Why might college graduates be more likely to respond to a survey than people with less education?

This also appears to be true in this survey too. In real life a lot more people are high school graduates or some college than college graduates, which is the opposite in this poll's results. College graduates are more likely to respond to a survey thant the other groups because they probably trust institutions more, and are more confident in answering questions about politics, civic life, the homefront etc.

## Task 3: Unweighted Voting Intentions by Education (4 points)

This is what the results look like if you treat every respondent equally — no weights applied.

```{r unweighted}
unweighted_by_education <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  group_by(educ, Q21) |>
  summarize(count = n(), .groups = "drop_last") |>
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

unweighted_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename("Yes (%)" = pct_1, "No (%)" = pct_2, "Unsure (%)" = pct_3)

unweighted_summary
```

**Question 3:** Based only on these unweighted numbers, write a one-sentence headline for a story about education and voting intentions.

More education and more likely to vote, an unweighted survey shows.

## Task 4: Weighted Voting Intentions by Education (2 points)

Now apply the weights. Instead of counting respondents, we sum their weights — so someone with a weight of 2.0 contributes twice as much as someone with a weight of 1.0.

```{r weighted}
weighted_by_education <- nonvoters_data |>
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  group_by(educ, Q21) |>
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

weighted_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename("Yes (%)" = pct_1, "No (%)" = pct_2, "Unsure (%)" = pct_3)

weighted_summary
```

## Task 5: Visualize the Difference (4 points)

```{r visualization}
viz_data <- bind_rows(
  unweighted_by_education |>
    filter(Q21 == 1) |>
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  weighted_by_education |>
    filter(Q21 == 1) |>
    mutate(Type = "Weighted",
           percentage = weighted_percentage) |>
    select(Type, educ, percentage)
)

ggplot(viz_data, aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
            position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  labs(
    title = "Weighted vs. Unweighted: Share Planning to Vote, by Education",
    x = "Education Level",
    y = "% Saying Yes"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#2166ac", "Weighted" = "#d6604d")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Question 4:** Which education group shows the biggest shift between weighted and unweighted results? In percentage points, how large is that shift? Does weighting change the story, or does it just change the exact numbers?

The education group that shows the biggest shift between weighted and unweighted results is the "High school or less" group. In percentage points the shift is 2.7%. The weighting dosen't change the overall story really, it just changes the exact numbers being used.

## Task 6: The UMD Belonging Survey (8 points)

Earlier in class this week, we looked at the 2024 UMD Belonging and Community Survey. That survey had a 23% response rate and the report acknowledged "statistically significant differences" between the sample and the UMD population — but applied no weighting to correct for them.

**Question 5:** Based on what you've learned in this assignment, write 2-3 paragraphs as if you were a reporter evaluating the UMD survey before writing a story based on it. Address: (1) what the lack of weighting means for how you'd use the results, (2) which groups in the UMD survey appeared most underrepresented based on Figures 1 and 2, and (3) what question you would ask the survey's authors before publishing.

The lack of weighting means that we can not use the results of the 2024 UMD Belonging and Community Survey. Due to the response rate being 23% and the statistically significant differences between the sample and the UMD population were not weighted to correct the sample to the population the sample data does not accurately reflect the UMD population. Without weighting the survey results, some student answers and groups are very over represented and others are very under represented.This also means that the results probably cater to the more extremes with the people most happy or mad about certain things, furthering that the data is not reliable.

Based on figures 1,2, and 3 the entire undergraduate population of students was under represented in the 2024 UMD survey. The survey's goal was to accurately reflect the entire UMD population, which is staff, faculty, grad and undergrad students and more. The majority of that population is undergraduate students, however the majority of the people who took the survey were staff and faculty. Without weighting those results it provides a while over coverage to the results of staff and faculty, and a massive under coverage of undergraduate students. Also, the demographics of the UMD population is roughly 50/50 male female. In the survey the sample demographics was 55% female and 38% male with another 7% reporting not knowing or gender expansive. Also roughly 12% of the people did not record race or ethnicity. This further provides evidence that the data from the poll the male students of color were probably completly underreperesentd in the survey. 

The question I have for the survey's authors is why did they chose not apply any weighting differences among the different demographic groups that asnwered to try and get an accurate representation of the population?


