---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```

### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
MandMs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```

### Task 3: Sample 5 rows of data from the dataset
```{r}
sampled_data <- MandMs |> slice_sample(n =5)
```

### Task 4: Calculate the mean age (just mean) and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
sampled_stats <- sampled_data |> summarize(mean_red = mean(red), sd_red = sd(red))
```
### Task 5: Calculate the mean age (just mean) and standard deviation of the population, using the same measure as in Task 4
```{r}
population_stats <- MandMs |> summarize(mean_red = mean(red), sd_red = sd(red))
```

### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The two sets of means are somewhat different from one another. The mean of the sample of the red M&M's is 84 while the mean of the population of red M&M's is 72.98. The sample is only 5 out of the 100, which is not a big sample. This could explain the reason why the mean is off the population, because it could have randomly sampled a couple of outlier collums. This is also shown in the how different the standard deviations are for the sample and population. The sample's standard deviation of red M&M's is 22.2 while the populations is 16.4.
### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

What this tells me is that small samples, like 5 out of 100, will lead to results that will greatly differ from the population. This tells me that multiple groups got outlier numbers throughout the M&M's dataset. However, the more random sampling is done the likelhood is that the random samples will eventually start becoming closer to the population. This essentially means the larger the random sample, the more it reflects the actual population.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
