---
title: "HW8b_PollingSituations"
output: html_document
---

---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")
```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}
ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)
```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

The value is roughly 0.105. This is much lower than the unafilliated proporition of about 0.36, which tells me that White Evangelicals make up a significantly less part of the population than people who are not affiliated with any religion.

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces |> sample_n(50)

proportion_50 <- mean(sample_50$evangelical)

```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

My sample estimate to the population proportion is relatively close. 
My sample estimate turned out to 0.16, which is only 0.06 (6 percent) off from the actual proportion, which from a sample of 50 out of 54803 is pretty close. This is also very similar to what I saw with the unafilliated group in class which I also got roughly 6 percent with the sample from the population proportion.

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}
set.seed(1966)

sample_500 <- ces |> sample_n(500)

proportion_500 <- mean(sample_500$evangelical)
```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

The larger sample did improve the estimate, the sample was at 0.09. This is now roughly only 1.04 percent off, much closer than the 6 percent. This one was even closer to the population than what I saw in class, but both follow the same pattern of being closer to the actual population.
### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
population_proportion <- mean(ces$evangelical)

simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share of White Evangelicals",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()

```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

I would say that the n=50 histogram is even more symmetric than the unaffiliiated histogram from class. This one seems even less spread out and somewhat even more clustered around the mean than the unaffiliated one is. I think this is the case because the unaffilaited group is much larger which could lead to more variance in data and samples then the White Evangelical group.

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

The distribution changes very drastically with a larger sample. The n=500 is much more clustered and less spread out than the n=50 sample. The improvement is similar, maybe only ever slightly smaller that what I saw in class with the unaffiliated group. That one was ever slightly more symmetrical, but at that point its semantics.

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

It would be about roughly 20 people, because the population proportion is roughly 10%, and ten percent on 200 is 20. This suggests that the reliability of estimates for smaller subgroups is not all that reliable because the samples are just not big enough. The small samples can have big outliers that skew data and misrepresent groups easily.

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

The first question is if the poll is randomly sampled. The next question I would have is how representative is the poll's sample of the population it's polling. If the rouglyh 40 people is a large enough sample of the population that would be good, but if not that would be very bad. If it isn't it would lead to data that could be very misleading for the actual population, and just plainly wrong. also, in a Presidential Election is the poll a nationwide poll, a state poll, or a local/municipal poll? This drastically changes the population, and whether this poll could or should be even used/done.
When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
