---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "YOUR NAME HERE"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
MocoRestaurants <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- MocoRestaurants |> summarize(mean_compliance_score = mean(compliance_score),
    median_compliance_score = median(compliance_score),
    sd_compliance_score = sd(compliance_score),
    min_compliance_score = min(compliance_score),
    max_compliance_score = max(compliance_score))


```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
mean_score <- mean(MocoRestaurants$compliance_score, na.rm = TRUE)

ggplot(MocoRestaurants, aes(x = compliance_score)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  geom_vline(aes(xintercept = mean_score), 
             color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of  MOCO Compliance Scores",
    x = "Compliance Score",
    y = "Number of Restaurants"
  ) +
  theme_minimal()

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

The histogram and standard deviation tell me that the distribution of compliance scores is somewhat clustered due to most of the compliance scores being within 1 standard deviation of the mean. Since the mean is at rougly 96, it falls within one standard deviation. This tells me that the distribution of the compliance score is roughly normal. Also just looking at the histogram you can tell most of the data points are around the mean.
---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
WMATARidership<- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- WMATARidership |> summarize(mean_bus = mean(bus),
    sd_bus = sd(bus),mean_rail = mean(rail), sd_rail = sd(rail)) 
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- WMATARidership |>
  mutate(
    rail_zscore = (rail - wmata_stats$mean_rail) / wmata_stats$sd_rail,
    bus_zscore  = (bus  - wmata_stats$mean_bus)  / wmata_stats$sd_bus
  )
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
wmata_zscores |>
 filter(abs(rail_zscore) > 2)
mean_rail <- mean(wmata_zscores$rail)
sd_rail <- sd(wmata_zscores$rail)


```
There are 6 unusual days. 
### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

The mean is off by over a thousand and so is the standard deviation. This suggests that random sampling can be incorrect and the more you do it, it becomes more reprasantitive of the population.
```{r}
set.seed(42)

wmata_sample <- WMATARidership |>
  sample_n(91)

wmata_sample |>
  summarise(
    mean_rail = mean(rail),
    sd_rail = sd(rail),
    mean_bus = mean(bus),
    sd_bus = sd(bus)
  )
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

The unusual scores in the z-score analysis suggest the story that random events and days such as bad weather could have really impacted the drop and ridership. Maybe holidays or sporting events.

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
CarThefts<- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- CarThefts |>
  mutate(theft_rate = (thefts / population) * 10000)
```
### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)


Given that the mean is 187 and the standard deviation is 42 the stastitcal measures that I would prioritze using is populaiton percentage versus the people going to the pools. I would first want to know the population of the area. Then I would want to know data of past years to compare to the data of this year. Next I would want to randomly sample from the population how many times they used the 12 community pools, and what percentege of the population does.

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

I would use Z-scores to identify how many standard deviations away each pool is from the mean attendance of the 12 community pools. This would show me how far away some pools are from  the average attendance, which are underpreforming, and which are overpreforming.

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would ask how many days are the pools open. Is the 60 days per pool a large enough sample for representing how many the days the pools are open. I would ask if all 12 community pools are being sampled. If they are not, then this would oversample one pool, and if the 60 days per pools sampling is not a big enough sample it could have data that wildy skews the overall data and misreperesnting the actual figures.
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
